A weakly structured stem for human origins in Africa

### 2023-01-09 paper.tex
Interglacial Discussion

The rapid rise in sea levels and increased precipitation during MIS
5e interglacial, following a glacial period of aridity across Africa
\citep{Blome2012-lw}, might have triggered migration inland away from the
coasts, as has been suggested, e.g., for the Paleo-Agulhas plain
\citep{Marean2014-pg}. Following these merger events, the stems subsequently fractured into
subpopulations which then persist over the past $\sim$120ka. These
subpopulations can be linked to contemporary groups despite subsequent gene
flow across the continent; for example, a genetic lineage sampled in the Gumuz
has a $0.70$ probability of being inherited from the ancestral ``eastern''
subpopulation $55$ka versus $0.06$ probability of being inherited
from the ``southern'' subpopulation (see Table~\ref{tab:supp-lineage-probabilitys} for
additional comparisons). We also find that Stem 2 continued to contribute to
western Africans during the Last Glacial Period, indicative that this gene flow
likely occurred in western/central Africa (Table~\ref{tab:migration-rates}).
Such an interpretation is reinforced by the differential migration rates
between regions, i.e. the gene flow from Stem 2 to western Africans is estimated to be 5x that
of the rate to eastern Africans during this period. We performed a variety of
validation tests to explore sensitivity of our assumptions, including relaxing
fixed parameters (SI Section~\ref{sec:validation_strategy}). Most validation
tests resulted in parameters similar to the models discussed above. However,
one exception was the inferred Out of Africa and eastern/western African
divergences which were 10-15ka younger than our fixed parameters. The complexity of
these events warrants the inclusion of additional Eurasian populations and
greater detail in modeling these population expansions; nonetheless, this model
mis-specification does not change other features of our inferred model.

>>>>

Shifts in wet/dry conditions across the African continent $140$ka to $100$ka may have
promoted these merger events between divergent stems. 
Precipitation does not neatly track interglacial cycles in Africa and heterogeneity across regions may mean 
that beginning of an arid period in eastern Africa is conversely the start of a wet period in southern Africa\citep{Blome2012-lw}.
The rapid rise in sea levels during the MIS 5e interglacial might have triggered migration inland away from the
coasts, as has been suggested, e.g., for the Paleo-Agulhas plain\citep{Marean2014-pg}. 
Following these merger events, the stems subsequently fractured into
subpopulations which then persist over the past $\sim$120ka. These
subpopulations can be linked to contemporary groups despite subsequent gene
flow across the continent; for example, a genetic lineage sampled in the Gumuz
has a $0.70$ probability of being inherited from the ancestral ``eastern''
subpopulation $55$ka versus $0.06$ probability of being inherited
from the ``southern'' subpopulation (see Table~\ref{tab:supp-lineage-probabilitys} for
additional comparisons).

We also find that Stem 2 continued to contribute to
western Africans during the Last Glacial Period, indicative that this gene flow
likely occurred in western/central Africa (Table~\ref{tab:migration-rates}).
Such an interpretation is reinforced by the differential migration rates
between regions, i.e. the gene flow from Stem 2 to western Africans is estimated to be 5x that
of the rate to eastern Africans during this period. We performed a variety of
validation tests to explore sensitivity of our assumptions, including relaxing
fixed parameters (SI Section~\ref{sec:validation_strategy}). Most validation
tests resulted in parameters similar to the models discussed above. However,
one exception was the inferred Out of Africa and eastern/western African
divergences which were 10-15ka younger than our fixed parameters. The complexity of
these events warrants the inclusion of additional Eurasian populations and
greater detail in modeling these population expansions; nonetheless, this model
mis-specification does not change other features of our inferred model.
